Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-10 submit/paper.tex
first commit

\maketitle
\begin{abstract}
  The notion of activity cliffs is an intuitive approach to characterizing structural features that
  play a key role in modulating biological activity of a molecule. A variety of methods have been
  described to quantitatively characterize activity cliffs, such as SALI and SARI. However, these
  methods are primarily retrospective in nature; highlighting cliffs that are already present in the
  dataset. The current study focuses on employing a pairwise characterization of a dataset to train
  a model to predict whether a new molecule will exhibit an activity cliff with one or more members
  of the dataset. The approach is based on predicting a value for pairs of objects rather than the
  individual objects themselves (and thus allows for robust models even for small SAR datasets). We
  extracted structure-activity data for several ChEMBL assays and developed random forest models to
  predict SALI values, from pairwise combinations of molecular descriptors. The models exhibited
  reasonable RMSE's though, surprisingly, performance on the more significant cliffs tended to be
  better than on the lesser ones. Our results indicate that the predictive models are able to
  prioritize molecules in terms of their ability to activity cliffs and can be used as a way to
  extend an observed activity landscape to identify molecules that could lead to significant
  improvements in activity. Conversely, the method can be used to deprioritize molecules that are
  predicted to show an activity cliff with a potent molecule already in the dataset.
\end{abstract}

@@

The landscape paradigm for SAR data was first proposed 20 years ago \cite{Johnson:1990ys} and has
recently seen a resurgence with a number of studies describing new ways to quantify and visualize
activity landscapes. When SAR data is viewed as a landscape, with the X-Y plane representing
structural characteristics (which will usually be a 2-dimensional representation of a
multi-dimensional descriptor space) and the Z-axis representing the observed activities, one can
identify two broad types of regions on the landscape - smooth rolling regions, corresponding to set
of molecules exhibiting continuous SAR (i.e., similar structures and similar activities) and rough,
gorge-like regions (i.e., very similar structures, but large differences in activity) corresponding
to molecules that exhibit SAR discontinuity. The latter have also been term activity cliffs
\cite{Maggiora:2006aa}. From a medicinal chemistry point of view, the latter regions of a landscape
can be the most interesting as they can provide insight into structural features that are key to
improving (or conversely reducing) potency. There is a rich history of methods that have correlated
structural differences with corresponding differences in activity -- matched molecular pairs
\cite{Leach:2006aa}, SAS maps \cite{Maggiora:2011kx} and more recently SALI \cite{Guha:2008aa} and
SARI \cite{Peltason:2007aa}. Both SALI and SARI focus on numerically characterizing a structure
activity landscape. The former is defined for a pair of molecules as
\begin{equation}
  \label{eq:1}
  S_{i,j} = \frac{|A_i - A_j|}{1 - sim(i,j)}
\end{equation}
where $A_i$ and $A_j$ represent the observed activities of molecules $i$ and $j$, and $sim(i,j)$
represents the structural similarity between the two molecules (usually based on some form of
fingerprint similarity). Using the SALI metric, one can take a collection of $n$ molecules and
represent them as an $n \times n$ matrix of SALI values - larger values representing more
significant activity cliffs. The SARI approach is based on a score defined as
\begin{equation}
  \label{eq:2}
  \mathrm{SARI} = \frac{1}{2} \left(\mathrm{score}_{cont} + (1 -
    \mathrm{score}_{disc}) \right)
\end{equation}
where the individual score terms are derived on the basis of potency and pairwise similarities. The
reader is referred to Ref. \citen{Peltason:2007aa} for a detailed discussion of this approach.

@@

In traditional QSAR modeling approaches, one simply predicts the activity of a new molecule and
would then evaluate the SALI (or SARI or some other measure) to determine whether the molecule leads
to an activity cliff. However, the fact that an activity cliff represents a SAR
discontinuity\cite{Maggiora:2006aa} implies that most statistical and machine learning methods will
be unlikely to predict very different activities for two structurally similar molecules. In other
words, a new molecule, similar to a subset of the training set, will tend to have a predicted value
that is similar to those molecules, rather than a drastically different value.

An alternative approach, that is the focus of this paper, is to directly predict SALI values for
pairs of molecules. Thus rather than predict individual activities, we predict SALI values for pairs
of molecules. This approach is somewhat similar to the SPREAD method \cite{Scheiber:2009fk} which
identified substructures that were predictive of activity differences. However, our solution
considers both activity differences and structural similarities. As a result, instead of ranking
compounds in terms of their predicted activity, we instead rank a compound in terms of its predicted
SALI; i.e., its predicted ability to exhibit an activity cliff when paired with other molecules in
the dataset. This approach could be useful when deciding how far to extend an analog series as well
as prioritizing scaffolds for further study.

@@

The paper is organized as follows. Section \ref{sec:datasets} describes the datasets used in this
study. Section \ref{sec:methodology} presents the methodology we employ to predict activity cliffs
and Section \ref{sec:applications} discusses the results of the predictive models. Finally, Sections
\ref{sec:discussion} and \ref{sec:conclusions} discusses some of the issues underlying this approach
and possible extensions of this work.

@@

\bibliography{paper}

>>>>

\maketitle
\begin{abstract}
  The notion of activity cliffs is an intuitive approach to
  characterizing structural features that play a key role in
  modulating biological activity of a molecule. A variety of methods
  have been described to quantitatively characterize activity cliffs,
  such as SALI and SARI. However, these methods are restrictive in the
  sense that they focus on exploratory analyses; highlighting cliffs
  that are already present in the dataset. The current study focuses
  on using activity cliffs in a dataset to train a model to predict
  whether a new molecule will exhibit an activity cliff with one or
  more members of the dataset. The approach is based on predicting
  pairs of objects rather than the individual objects themselves (and
  thus allows for robust models even for small SAR datasets). We
  extracted structure-activity data for several ChEMBL assays and
  developed random forest models to predict SALI values, from pairwise
  combinations of molecular descriptors. The models exhibited
  reasonable RMSE's though, surprisingly, performance on the more
  significant cliffs tended to be better than on the lesser ones. Our
  results indicate that the predictive models are able to prioritize
  molecules in terms of their ability to activity cliffs and can be
  used as a way to extend an observed activity landscape to identify
  molecules that could lead to significant improvements in
  activity. Conversely, the method can be used to deprioritize
  molecules that are predicted to show an activity cliff with a potent
  molecule already in the dataset.
\end{abstract}

@@

The landscape paradigm for SAR data was first proposed 20 years ago
\cite{Johnson:1990ys} and recently has seen a resurgence with a number
of studies describing new ways to quantify and visualize activity
landscapes. When SAR data is viewed as a landscape, with the X-Y plane
representing structural characteristics (which will usually be a
2-dimensional representation of a multi-dimensional descriptor space)
and the Z-axis representing the observed activities, one can identify
two broad types of regions on the landscape - smooth rolling regions,
corresponding to set of molecules exhibiting continuous SAR (i.e.,
similar structures and similar activties) and rough, gorge-like
regions (i.e., very similar structures, but large differences in
activity) corresponding to molecules that exhibit SAR
discontinuity. The latter have also been term activity cliffs
\cite{Maggiora:2006aa}. From a medicinal chemistry point of view, the
latter regions of a landscape can be the most interesting as they can
provide insight into structural features that are key to improving (or
conversely reducing) potency. There is a rich history of methods that
have correlated structural differences with corresponding differences
in activity - matched molecular pairs \cite{Leach:2006aa}, SAS maps
\cite{Maggiora:2011kx} and more recently SALI \cite{Guha:2008aa}
and SARI \cite{Peltason:2007aa}. Both SALI and SARI focus on
numerically characterizing a structure activity landscape. The former
is defined for a pair of molecules as
\begin{equation}
  \label{eq:1}
  S_{i,j} = \frac{|A_i - A_j|}{1 - sim(i,j)}
\end{equation}
where $A_i$ and $A_j$ represent the observed activities of molecules
$i$ and $j$, and $sim(i,j)$ represents the structural similarity
between the two molecules (usually based on some form of fingerprint
similarity). Using the SALI metric, one can take a collection of $n$
molecules and represent them as an $n \times n$ matrix of SALI values
- larger values representing more significant activity cliffs. The
SARI approach is based on a score defined as 
\begin{equation}
  \label{eq:2}
  \mathrm{SARI} = \frac{1}{2} \left(\mathrm{score}_{cont} + (1 -
    \mathrm{score}_{disc}) \right)
\end{equation}
where the individual score terms are derived on the basis of potency
and pairwise similarities. The reader is referred to
Ref. \citen{Peltason:2007aa} for a detailed discussion of this
approach.

@@

In traditional QSAR modeling approaches, one simply predicts the
activity of a new molecule and would then evaluate the SALI (or SARI
or some other measure) to determine whether the molecule leads to an
activity cliff. However, the fact that an activty cliff represents a
SAR discontinuity\cite{Maggiora:2006aa} implies that most statistical
and machine learning methods will be unlikely to predict very
different activities for two structurally similar molecules. In other
words, a new molecule, similar to a subset of the training set, will
tend to have a predicted value that is similar to those molecules,
rather than a drastically different value.

An alternative approach, that is the focus of this paper, is to
directly predict SALI values for pairs of molecules. Thus rather than
predict individual actvities, we predict SALI values for pairs of
molecules. This approach is somewhat similar to the SPREAD method
\cite{Scheiber:2009fk} which identifed substructural features that
were predictive of activity differences. However, our solution
considers both activity differences and structural similarities. As a
result, instead of ranking compounds in terms of their predicted
activity, we instead rank a compound in terms of its predicted SALI;
i.e., its predicted ability to exhibit an actvity cliff when paired
with other molecules in the dataset. This approach could be useful
when deciding how far to extend an analog series as well as
prioritizing scaffolds for further study.

@@

The paper is organized as follows. Section \ref{sec:datasets}
describes the datasets used in this study. Section
\ref{sec:methodology} presents the methodology we employ to predict
activity cliffs and Section \ref{sec:applications} discusses the
results of the predictive models. Finally, Sections
\ref{sec:discussion} and \ref{sec:conclusions} dicusses some of the
issues underlying this approach and possible extensions of this work.

@@

%%\bibliography{paper}
\providecommand{\refin}[1]{\\ \textbf{Referenced in:} #1}
\begin{thebibliography}{10}

\bibitem{Johnson:1990ys}
Johnson,~M.;\ \ Maggiora,~G. \textit{Concepts and Applications of Molecular
  Similarity;} John Wiley \& Sons: New York, 1990.

\bibitem{Maggiora:2006aa}
Maggiora,~G.~M.  On Outliers and Activity Cliffs--Why {QSAR} Often
  Disappoints..  \textit{J.~Chem.~Inf.~Model.} \textbf{2006,} \textsl{46,}
  1535--1535.

\bibitem{Leach:2006aa}
Leach,~A.;\ \ Jones,~H.;\ \ Cosgrove,~D.;\ \ Kenny,~P.;\ \ Ruston,~L.;\ \
  MacFaul,~P.;\ \ Wood,~J.;\ \ Colclough,~N.;\ \ Law,~B.  Matched Molecular
  Pairs as a Guide in the Optimization of Pharmaceutical Properties; a Study of
  Aqueous Solubility, Plasma Protein Binding and Oral Exposure.
  \textit{J.~Med.~Chem.} \textbf{2006,} \textsl{49,} 6672--6682.

\bibitem{Maggiora:2011kx}
Shanmugasundaram,~V.;\ \ Maggiora,~G.  Characterizing Property and Activity
  Landscapes Using an Information-Theoretic Approach.   In  \textit{CINF-032.
  222nd ACS National Meeting, Chicago, IL, United States}; American Chemical
  Society: Washington, D.C., 2001.

\bibitem{Guha:2008aa}
Guha,~R.;\ \ Van~Drie,~J.  The Structure-Activity Landscape Index:
  {I}dentifying and Quantifying Activity-Cliffs.  \textit{J.~Chem.~Inf.~Model.}
  \textbf{2008,} \textsl{48,} 646--658.

\bibitem{Peltason:2007aa}
Peltason,~L.;\ \ Bajorath,~J.  SAR Index: Quantifying the Nature of
  Structure-Activity Relationships.  \textit{J.~Med.~Chem.} \textbf{2007,}
  \textsl{50,} 5571--5578.

\bibitem{Wawer:2008aa}
Wawer,~M.;\ \ Peltason,~L.;\ \ Weskamp,~N.;\ \ Teckentrup,~A.;\ \ Bajorath,~J.
  Structure-Activity Relationship Anatomy by Network-like Similarity Graphs and
  Local Structure-Activity Relationship Indices.  \textit{J.~Med.~Chem.}
  \textbf{2008,} \textsl{51,} 6075--6084.

\bibitem{Vogt:2011bs}
Vogt,~M.;\ \ Huang,~Y.;\ \ Bajorath,~J.  From Activity Cliffs to Activity
  Ridges: {I}nformative Data Structures for SAR Analysis.
  \textit{J.~Chem.~Inf.~Model.} \textbf{2011,} \textsl{51,} 1848--1856.

\bibitem{Dimova:2011fk}
Dimova,~D.;\ \ Wawer,~M.;\ \ Wassermann,~A.~M.;\ \ Bajorath,~J.  Design of
  Multitarget Activity Landscapes That Capture Hierarchical Activity Cliff
  Distributions.  \textit{J.~Chem.~Inf.~Model.} \textbf{2011,} \textsl{51,}
  258--266.

\bibitem{Wawer:2010fv}
Wawer,~M.;\ \ Bajorath,~J.  Similarity-potency trees: {A} method to search for
  {SAR} information in compound data sets and derive {SAR} rules.
  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,} \textsl{50,} 1395--1409.

\bibitem{Hu:2010kl}
Hu,~Y.;\ \ Bajorath,~J.  Molecular Scaffolds with High Propensity to Form
  Multi-Target Activity Cliffs.  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,}
  \textsl{50,} 500--510.

\bibitem{Wassermann:2010oq}
Wassermann,~A.~M.;\ \ Bajorath,~J.  Chemical Substitutions that Introduce
  Activity Cliffs Across Different Compound Classes and Biological Targets.
  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,} \textsl{50,} 1248-1256.

\bibitem{Peltason:2009vn}
Peltason,~L.;\ \ Weskamp,~N.;\ \ Teckentrup,~A.;\ \ Bajorath,~J.  Exploration
  of Structure-Activity Relationship Determinants in Analogue Series.
  \textit{J.~Med.~Chem.} \textbf{2009,} \textsl{52,} 3212--3224.

\bibitem{Scheiber:2009fk}
Scheiber,~J.;\ \ Jenkins,~J.~L.;\ \ Bender,~A.;\ \ Milik,~M.;\ \ Mikhalov,~D.;\
  \ Sukuru,~S.;\ \ Cornett,~B.;\ \ Whitebread,~S.;\ \ Laszlo,~U.;\ \
  Davies,~J.;\ \ Glick,~M.  {SPREAD} - exploiting chemical features that cause
  differential activity behavior.  \textit{Stat.~Anal.~Data Mining}
  \textbf{2009,} \textsl{2,} 115-122.

\bibitem{Cavalli:2002aa}
Cavalli,~A.;\ \ Poluzzi,~E.;\ \ De~Ponti,~F.;\ \ Recanatini,~M.  Toward a
  Pharmacophore for Drugs Inducing the Long {QT} Syndrome: Insights from a
  {CoMFA} Study of {HERG K+} Channel Blockers.  \textit{J.~Med.~Chem.}
  \textbf{2002,} \textsl{45,} 3844--3853.

\bibitem{Costanzo:2005uq}
Costanzo,~M.~J.;\ \ Almond,~Jr,~H.~R.;\ \ Hecker,~L.~R.;\ \ Schott,~M.~R.;\ \
  Yabut,~S.~C.;\ \ Zhang,~H.-C.;\ \ Andrade-Gordon,~P.;\ \ Corcoran,~T.~W.;\ \
  Giardino,~E.~C.;\ \ Kauffman,~J.~A.;\ \ Lewis,~J.~M.;\ \ de~Garavilla,~L.;\ \
  Haertlein,~B.~J.;\ \ Maryanoff,~B.~E.  In-depth study of tripeptide-based
  alpha-ketoheterocycles as inhibitors of thrombin. {E}ffective utilization of
  the {S1'} subsite and its implications to structure-based drug design.
  \textit{J.~Med.~Chem.} \textbf{2005,} \textsl{48,} 1984--2008.

\bibitem{Kalla:2006kx}
Kalla,~R.~V.;\ \ Elzein,~E.;\ \ Perry,~T.;\ \ Li,~X.;\ \ Palle,~V.;\ \
  Varkhedkar,~V.;\ \ Gimbel,~A.;\ \ Maa,~T.;\ \ Zeng,~D.;\ \ Zablocki,~J.
  Novel 1,3-disubstituted 8-(1-benzyl-1H-pyrazol-4-yl) xanthines: high affinity
  and selective A2B adenosine receptor antagonists.  \textit{J.~Med.~Chem.}
  \textbf{2006,} \textsl{49,} 3682--3692.

\bibitem{Dai:2007kl}
Dai,~Y. \textit{et al.}\   Discovery of
  N-(4-(3-amino-1H-indazol-4-yl)phenyl)-N'-(2-fluoro-5-methylphenyl)urea
  (ABT-869), a 3-aminoindazole-based orally active multitargeted receptor
  tyrosine kinase inhibitor.  \textit{J.~Med.~Chem.} \textbf{2007,}
  \textsl{50,} 1584--1597.

\bibitem{Breiman:1984aa}
Breiman,~L.;\ \ Friedman,~J.;\ \ Olshen,~R.;\ \ Stone,~C.
  \textit{Classification and Regression Trees;} Chapman \& Hall/CRC: Boca
  Raton, FL, 1984.

\bibitem{Breiman:2001nx}
Breiman,~L.  Statistical Modeling: Two Cultures.  \textit{Stat.~Sci.}
  \textbf{2001,} \textsl{16,} 199--231.

\bibitem{r}
{R Development Core Team}, ``{R}: {A} Language and Environment for Statistical
  Computing'',  R Foundation for Statistical Computing, Vienna, Austria, 2008
  {ISBN} 3-900051-07-0.

\end{thebibliography}
